---
title: "Final Report: The Effects of Neutering on Female and Male Adult Cats"
author: "Chelsea Noack"
date: "December 18, 2017"
output:
    github_document: default
    word_document: default
bibliography: references.bib
csl: bioinformatics.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

The affection for fat cats is an affliction. As a culture, we glorify at cats which are overweight or obese. For what we often mistake as chubby cuteness is a danger to the animals. Fat cats are at risk for diseases such as diabetes type 2, lameness (a dysfunction in the locomotor system), oral cavity disease, urinary tract disease, and neoplasia (abnormal tissue growth [@takosova2016feline]. Diabetes type 2 leads to less insulin which balances blood, sugar, or glucose levels. Lameness, while not considered an illness, makes a cat's life more difficult by needing to limp. Oral cavity disease is an inflammatory and ultimately debilitating sickness. Urinary tract disease encompasses many problems for a cat's urethra or bladder. Lastly, neoplasia can be a non-cancerous or tumorous growth, both undesirable. The life of a fat cat becomes hindered and limited in basic day-to-day activities as well. Being obese makes it more difficult to fit in tight spaces, run, jump, and breathe. The list is unfortunately long. 

In the United States, we find that obesity in felines is one of the most predominant nutritional disorders. Between 11.5% and 63% of cats being overweight or obese [@tarkosova2016feline]. So, what causes obesity in felines? 

The causes of obesity are difficult to pinpoint. Diet is a huge factor. Owners often over-estimate how much a cat needs to eat. Or a cat will have an increased appetite due to hormones. Genetics also plays a factor. Like humans, cats can inherit slow metabolisms, making digestion slower. Even molecular factors. For example, FT4, free serum throxine, has a specific range within the thyroid which is optimal. When FT4 increases outside the range, chances of obesity increases. This occurs in both humans and felines [@ferguson2007obesity]. With the many causes of obesity, it is difficult to pinpoint *how* best to prevent feline obesity. 

Yet, one other factor remains, and that is neutering.

Many studies point to a relationship between neutering and weight gain. A study done by Kanchuck et al. attempts to find out what the cause of post-neutering obesity was. They observe two possibilities, either increased food intake or less energy expenditure. They find that after neutering, cats increase their appetite and food intake [@kanchuk2003weight]. Decreased energy expenditure did not play a statistically significant role. And while most studies find a relationship, they do not pinpoint *how* neutering plays a role.

Yet, neutering is important for population control, conservation, feline health and human health. 

Humans and felines have shared the same space for over 9,000 years. But, felines in an urban population is a health hazard [@mendes2011reduction]. 

Feral cats often dominate urban settings. They are outdoor cats with little to no human contact. Currently, there are approximately 50-80 million feral cats in the United States [@ireland2016spatial]. Reports estimate 80% of kittens born each year come from feral cats [@ireland2016spatial]. Feral cats are generalist predators, meaning they have a broad diet, they kill a lot of wildlife. Thus, they can be an invasive species. For example, on islands, cats contribute to 14% of bird, mammal, and reptile extinctions. They also contribute to 8% of critically endangered birds, mammals, and reptiles [@medina2011global]. 

Along with to conservation issues, neutering promotes health for cats and humans. It can prevent diseases such as reproductive tract disease, testicular neoplasia, and hormone-associated disorders like vaginal prolapse [@kustritz2012effects]. Neutering can prolong a cats life and prevent cancers, especially breast and testicular. It can also prevent female cats from going into heat. This is good, as when a female cat goes into heat, they will cry and urinate more than usual for 4-5 days for three weeks. While it's a natural biological process, it is not desirable for a cat living indoors. Feral cats not under population control can spread disease to humans as well. Such diseases like rabies and toxoplasmosis are the most common [@ireland2016spatial]. 

Thus, the topic becomes complicated. 

Neutering is an effective and common way to tackle population control. Lack of population control contributes to conservation issues, human health, and feline health. This is why veterinarians and institutions such as the A.S.P.C.A (American Society for the Prevention of Cruelty to Animals) advise neutering cats. But, studies indicate that neutering also can lead to obesity. And as shown, obesity affects the health of the cat. The conundrum isn't clear. The only way to better clarify the issue is more data analysis from previous studies. As scientists whom care for animals, we have a responsibility to find out more about neutering and its long-term effects on feline health. In doing so, we can better understand the advantages or disadvantages after neutering. It can also inspire alternative methods for population control.

My analysis attempts to use research done in a previous study to understand neutering's effects. I use data from a study titled *Effects of obesity, energy restriction and neutering on the faecal microbiota of cats* by Manuela M. Fischer, et al. (2017). 

The study focuses on both obese and lean cats using their fecal DNA in a controlled environment. I hypothesize that first, neutering increases body mass. Second, that females will be more obese, especially after neutering. I base this hypothesis from what we already know about human females. The sex hormone, estrogen, in females contributes to weight gain [@grantham2014estrogen]. Because cats are companion animals to humans in the scientific community, I assume female cats have a similar mechanism. To further support my analysis, I will also look at the gut microbiome in the cats and the age of the cats. I argue that sex will play a bigger role when understanding neutering rather than gut microbiota or age. 

# Methods

## Study design

Fischer et al. used twenty-four cats from the University of California, Davis with a median age of 6.4 years. All cats were pathogen free domestic shorthairs. From the twenty-four cats, eight of the cats were obese with four males and four females. Eight cats were lean and not neutered, again with four males and four females. Eight cats, six males and two females, were lean and neutered [@fischer2017effects]. All twenty-four cats had the same diet before and during the study. Every cat underwent a physical and blood sample before the experiment began.

Cats which were lean and both neutered or not neutered housed together. All had the same diet *but* their intake was not measured. The night before the fecal sampling, the cats slept alone. Fischer et al. performed BCS (body composition scores) for each cat before sampling. 

The obese, neutered cats underwent castration/spaying 1-6 years before the study [@fischer2017effects]. Unlike the lean cats, the obese cats slept alone. The obese neutered ate the normal diet given to all cats for ten days. For six weeks, they ate 60-70% of the prior intake. Again, unlike the lean cats, their intake was now measured. Fischer et al. recorded the cats' weights every week. They also recorded fecal samples before and after the dietary restriction began [@fischer2017effects].

## Sample origin and sequencing

Before sequencing the data, Fischer et al. collected fecal samples from the cat litter box once a day for three days straight. They monitored the cats and they did not collect any samples which were 15 minutes old. To get the bacterial DNA, Fischer et al. used bead-beating methods. Bead-beating is a way to release DNA, RNA, and proteins in the cells of a sample. They used a commercial DNA extraction kit [@fischer2017effects]. To bead-beat, samples go in tubes then on a homogenizer. A homogenizer which is a lab tool used to mix materials using high energy. They mixed "for 60 seconds at a speed of 4 m/s" [@fischer2017effects].

The 16S rRNA amplified using a universal bacterial primer. A universal bacterial primers is important to interpret PCR results. Especially for microbial communities [@mao2012coverage]. PCR (polymerase chain reaction) is a technique used to make a lot of copies of a segment of DNA. 

After PCR amplification, Fischer et al. added amplicon products of different samples. They purified the samples using AgencourtAmpure beads. The beads were important for high-throughput purification of the PCR amplicons. Then the samples combined in equal concentrations. Fischer et al. sequenced th samples using Roche 454 FLX titanium instruments [@fischer2017effects]. Pyrosequencing using instruments like Roche 454 have big advantages versus traditional Sanger sequencing. Primarily because they can examine many samples from many microbial communities [@tamaki2011analysis]. Once the raw sequences were complete, I downloaded them for my analysis. I downloaded them from NCBI's Sequence Read Archive RP066010. 

## Computational

First, I downloaded the data from the provided link. This was a two-part process. First, I had to download the SRAtoolkit from NCBI's website. Because the data I used was in .sra format, the toolkit converts it to .fastq. I wanted .fastq so I could perform a quality check later on. After downloading the toolkit, I performed a check to make sure it worked. The steps taken came from the toolkit's website. Once I assured that it worked, I wrote a Bash script which downloaded the data into my computer. Because the files were fairly small, this did not take much time. The files downloaded into my raw data folder and saved as fastq, thanks to the toolkit. 

With the data downloaded into my computer, I transferred it to RStudio to begin my analysis. In order to do so, I downloaded necessary, but helpful, libraries. The libraries included dplyr, tidyr, knitr, ggplot2, citr, seqinr, mctoolsr, dada2, and phyloseq. Dplyr, tidyr, knitr, and ggplot2 are general use packages to help analyze the data. 

DADA2 and phyloseq were extremely important for this process. DADA2 comes from bioconductor. Bioconductor serves as the primary package to clean up and analyze our data. DADA2 specifically cleans up amplicon errors [@callahan2016]. It can also extract sample names from their original .fastq names. Phyloseq takes on microbiome analysis in R [@mcmurdie2013]. Once all files downloaded and organized, I did a QC (quality check) report on each individual file. 

Initially, I trimmed the files based off of only the QC scores (they dropped around ~400 bp). After later reviewing the amount of errors in the sequences, I decided to extend the trim to 600. To counterbalance this, I made sure the quality cut off was at a smaller number (originally at 5, now at 2). Thus, we would get more data to analyze with adequate quality. After a proper trim and error check, I deleted any duplicate sequences. 454 Roche technology sequenced the data, so I used the suggested parameters based off of DADA2's guidelines. This is when DADA2 proved to be vital, as it facilitated the process for my specific amplicon data set. In the final steps, I observed the distribution of trimmed and denoised sequences. While the data was poorly sequenced, for my purposes it would work alright. Then, I removed chimeras. 

Now, the data given is fully processed. Before analysis, I created a metadata table and a melted table to observe during research. 

# Results

I begin my analysis by showing a general outlook of the data. A grid plot shows the abundance of  the sample groups (lean intact, lean neutered, obese neutered before weight loss and after weight loss) in Figure 1. We see neutered males are in the highest abundance. Females are in less abundance, with the highest being obese after energy restriction. The least abundant of the data is males who are intact, obese males, and females who are intact. 

Sometimes, a density plot can give a better idea of abundance than a grid plot. In a density plot, I plot overall abundance to total body mass (Figure 2). While the plot does not include sex of the cats, it indicates a trend. Generally, cats obese undergoing weight loss are the most dense. The second-most dense were the obese cats. This was close to findings in Figure 1. Lean, intact cats within the sampling distribution are overall in less abundance. They show a uniform distribution in mass. Neutered cats show the greatest range of total body mass (Figure 2). 

In sum, males are the most dominant in the samples. Especially the obese cats undergoing energy restriction. Females were the least dominant in the samples. Especially lean, intact female cats. 

Next, I decided to look explicitly into body mass in the cats. 

**The role of body mass**

I first asked what were the mean body masses among the four phenotypes for both male and female cats? 

Phenotypes against mean body masses and sex were put into a table (Table1). The phenotype with the highest mean body mass are the obese male cats. The second highest mean body mass were the obese male cats undergoing weight loss. The phenotype with the lowest mean body mass are the neutered female cats, followed by intact males. The medians of body mass are obese male cats and neutered female cats. The information in the table can be put on a graph for better interpretation, so that's what I did. 

We do not see differences between females and males in terms of patterns of body mass to phenotype (Figure 3). Male and female cats tend to follow a similar pattern. The largest difference, if any, is between intact females and males. The obese cats undergoing weight loss are nearly similar between males and females.  With this in mind, males and females generally follow a similar trend of increased body mass when neutered vs. intact. But, males will be generally larger than females within every phenotype (Figure 3). 

In sum, male cats were on average larger than female cats. The lightest phenotypes were intact, not neutered. 

Following my analysis of body mass, I determined to look for support from the microbial community found in the cat's gut. 

**The role of microbial communities**

Two density plots show which microbial communities appear in phenotype and sex (Figure 4 and Figure 5). The two plots are either the lower and upper weight ranges for the cats. I divide the plots for better visualization of the data as there were many data points to take into account. We see in the lower range of body masses, female cats dominate among all phenotypes. They show a fairly normal distribution (Figure 4). The phylum Firmicutes is the most abundant in the lower weight ranges for both males and females. Males dominate the upper body weight range and have two uneven peaks. In the upper body weight range, females have a moderate appearance with two peaks (Figure 5). Again, Firmicutes have the highest density of both males and females. In both plots, the phylum Bacteriodetes appears in both females and males. We also see phyla Proteobacteria and Fusobacteria fluctuate between sex and body mass (Figure 5). 

The microbial community question continued to linger, so I dug further. 

A grid plot is used for better visualization of my question. A new category called not applicable (N/A) occurs the most between females and males (Figure 6). After not applicable, Firmicutes, Bacteriodetes, and Actinobacteria occur the most. The phyla which occur least are Fusobacteria, Proteobacteria, and Tenericutes.

In sum, the dominating microbial phyla are unknown phylas, Firmicutes, Bacteriodetes, and Actinobacteria. The least dominant phyla are Fusobacteria, Proteobacteria, and Tenericutes. Females and males do not differ in microbial communities given the data.

Finally, I used the analysis of age as the final support for my hypothesis. 

**The role of age**

I asked whether age is a factor between sex and remaining obese after neutering. 

In a table, I put mean ages across phenotypes (Table 2). Neutered females and female obese cats (before and after weight loss) are the oldest of the sample. Neutered males, intact males, and obese males are the youngest of the sample (Table 2). The table is then put into a figure for visualization. Females appear to have a larger mean age and males dominate the younger age groups (Figure 7). Intact cats all have fairly similar age patterns for both males and females. Neutered obese before and after energy restriction have the most age variance between males and females (Figure 7). 

In order to weave everything together, mean age is put against bacterial phylas for females and males (Table 3). It's a hefty task. We see females still dominate, on average, the older age groups. Compared to other phenotypes, intact lean cats have considerably less bacteria phyla. The dominating theme is that females are, on average, still older than the males. 

Visualizing the hefty amount of data from the table could be many different plots. I chose bar plot as the most effective solution for visualization. We see that the intact lean phenotype will have three less bacterial phyla than obese or lean neutered phenotypes (Figure 8). On average, the neutered phenotypes of both lean and obese cats have older ages. Intact phenotypes are typically younger. The obese cats, both before and after energy restriction, have similar ages and similar bacterial phylas. Neutered phenotypes have older ages, on average. Obese and obese undergoing weight loss typically have similar ages as well as bacterial phylas. 

In sum, neutered obese female cats are older and male intact cats are youngest of the sample. Intact cats have three less bacterial phyla than the neutered phenotypes. Neutered, either lean or obese, was eldest of the sampling overall. Obese cats have similar ages and bacterial phyla.

```{r load-libraries, echo=FALSE, message = FALSE}
# Be sure to install these packages before running this script
# They can be installed either with the intall.packages() function
# or with the 'Packages' pane in RStudio

# load general-use packages
library("dplyr")
library("tidyr")
library("knitr")
library("ggplot2")

# this package allows for the easy inclusion of literature citations in our Rmd
# more info here: https://github.com/crsh/citr
# and here:
# http://rmarkdown.rstudio.com/authoring_bibliographies_and_citations.html
library("citr")

# These are the primary packages well use to clean and analyze the data
# this package needs to be installed from bioconductor -- it's not on CRAN
# see info here: https://benjjneb.github.io/dada2/dada-installation.html
library("dada2")

# This to export a fasta of our final denoised sequence variants
library("seqinr")

# To install this you have to install from GitHub
# See more info here: https://github.com/leffj/mctoolsr
# run this -- install.packages("devtools")
# and then this -- devtools::install_github("leffj/mctoolsr")
library("mctoolsr")

# And this to visualize our results
# it also needs to be installed from bioconductor
library("phyloseq")

# Install devtools package
devtools::install_github("ropenscilabs/gramr")
```

```{r phyloseq-object, echo = FALSE}
load("output/phyloseqobject.RData")
```

```{r meltedtable, echo = FALSE}
# Compile into a table
melted_obj <- psmelt(phyloseq_obj)
```

```{r gridplot-sex-abundance-phenotype}
# First figure indicating the general schema of the data
ggplot(data = melted_obj,
          aes(x = host_sex_s, 
              y = Abundance)) +
  geom_point() +
  facet_grid(. ~ host_phenotype_s) +
  labs(title = "Abundance of male and female cats with different phenotypes",
       x = "Sex of Cat")
```

**Figure 1:** A grid plot of abundance of female and male cats whom are intact, neutered, obese, and who were obese before weight loss. 

```{r density-mass-phenotype}
# Second figure using density to show the range of body mass to phenotype
melted_obj %>%
ggplot(aes(host_tot_mass_s, fill = host_phenotype_s)) +
  geom_density(alpha = 0.1) +
  xlim(3400, 7000) +
  labs(title = "Distribution of body size in different phenotypes",
       x = "Total Mass of Cats",
       y = "Density",
       fill = "Cat Phenotype")
```

**Figure 2**: A density plot of host's total mass to its phenotype. 

```{r tableone, echo = FALSE}
# Comment here
meanmass_8 <- melted_obj %>%
  group_by(host_phenotype_s, host_sex_s) %>%
  summarize(mean_bodymass = mean(host_tot_mass_s)) %>%
  arrange(desc(host_phenotype_s)) %>%
  head(n = 8)

# then we can use the `kable()` function to make a nicely formatted
# markdown table
meanmass_8 %>%
  kable()
```

**Table 1**: Mean body mass among phenotypes and sexes. 

```{r sex-mass-phenotype-barplot}
meanmass_8 %>%
  ggplot(aes(x = host_sex_s,
             y = mean_bodymass,
             fill = host_phenotype_s)) +
    geom_col(position = "dodge") +
  ggtitle("Cat sex to mean body mass over four phenotypes") +
  xlab("Cat Sex") + 
  ylab("Mean Body Mass") + 
  labs(fill = "Cat Phenotype")
```

**Figure 3**: Using the table, the mean body mass visualized with a bar graph. 

```{r density-sex-lowermass-phylum}
# Comment here
melted_obj %>%
ggplot(aes(host_tot_mass_s, fill = host_sex_s, colour = Phylum)) +
  geom_density(alpha = 0.1) +
  xlim(2500, 5000) +
  ggtitle("Lower range body mass to bacterial phyla and cat sex") +
  xlab("Cat total mass") +
  ylab("Density") +
  labs(fill = "Cat sex", colour = "Bacteria phylum")
```

**Figure 4**: A density plot of the lower range weight class of females and males and the correlating Phylums. 

```{r density-sex-uppermass-pylum}
# Comment here
melted_obj %>%
ggplot(aes(host_tot_mass_s, fill = host_sex_s, colour = Phylum)) +
  geom_density(alpha = 0.1) +
  xlim(5500, 7000) +
    ggtitle("Upper range body mass to bacterial phyla and cat sex") +
  xlab("Cat total mass") +
  ylab("Density") +
  labs(fill = "Cat sex", colour = "Bacteria phylum")
```

**Figure 5**: A density plot of the upper range weight class of females and males and the correlating Phylums. 

```{r gridplot-abundance-sex-phylum}
# Comment here
ggplot(data = melted_obj,
          aes(x = host_sex_s, y = Abundance)) +
  geom_point() +
  facet_grid(. ~ Phylum) +
  labs(title = "Abundance of male and female cats to bacterial phylum",
       x = "Sex of Cat")
```

**Figure 6**: A grid plot of the abundance of bacterial phylums in both males and females. 

```{r table2, echo = FALSE}
# Comment here
meanage_10 <- melted_obj %>%
  group_by(host_phenotype_s, host_sex_s) %>%
  summarize(mean_age = mean(host_age_s)) %>%
  arrange(desc(host_phenotype_s)) %>%
  head(n = 10)

# then we can use the `kable()` function to make a nicely formatted
# markdown table
meanage_10 %>%
  kable()
```

**Table 2**: A table of the mean ages for each phenotype in male and female cats. 

```{r baplot-phenotype-sex-age}
# Comment here
meanage_10 %>%
  ggplot(aes(x = host_phenotype_s,
             y = mean_age,
             fill = host_sex_s)) +
    geom_col(position = "dodge") +
  labs(title = "Male and female phenotypes measured to mean age",
       x = "Cat Phenotype",
       y = "Cat Age",
       fill = "Cat Sex")
```

**Figure 7**: A bar plot indicating different mean ages between the categories intact, neutered, obese, and obese after weight loss among females and males. 

```{r table3, echo = FALSE}
# Comment here
meanaps_30 <- melted_obj %>%
  group_by(host_phenotype_s, Phylum, host_sex_s) %>%
  summarize(mean_agephylumsex = mean(host_age_s)) %>%
  arrange(desc(host_phenotype_s)) %>%
  head(n = 50)

# then we can use the `kable()` function to make a nicely formatted
# markdown table
meanaps_30 %>%
  kable()
```

**Table 3**: A table indicating mean ages for the different cat phenotypes, bacteria phylas, and sex. 

```{r barplot-age-phenotype-phylum}
# Comment here
meanaps_30 %>%
  ggplot(aes(x = host_phenotype_s,
             y = mean_agephylumsex,
             fill = Phylum)) +
    geom_col(position = "dodge") +
    labs(title = "Phenotypes and bacterial phyla measured to mean age",
       x = "Cat Phenotype",
       y = "Cat Age",
       fill = "Phylum")
```

**Figure 8**: A bar plot showing age against phylum and category of intact, neutered, obese, and obese after weight loss. 

# Discussion

My hypothesis was that female cats are more obese than male cats, especially after being neutered. However, the data indicates opposite results. Male cats are heavier than female cats within all phenotypes; intact, neutered, obese, and obese with energy restriction (Table 1). The findings are supported by supplementary studies indicating that male cats often gain weight after being neutered. Specifically, neutering male cats in the earlier stages (between weeks 3-10) or in the late stages (weeks 34-41) of sexual development can alter feeding behaviors [@allaway2016metabolic].

First, I contend that there is a slight skew in the data when it comes to the abundance of male cats versus female cats, as shown in Figure 1. Arguably, the skew is too slight to make a difference. However, clearly there is some reason as to why male cats are heavier than the female cats. Thus, rather than using the sex to determine abundance, using the density of a particular phenotype to total body mass shows a different story. 

**Male cats are more obese than female cats**

We found that those who were obese undergoing weight loss had the thickest density of sequences in our data. I found this interesting as there were more lean cats, both intact and neutered, than the eight obese cats in total. In fact, the intact lean cats generally had a uniform distribution of total mass. Neutered lean cats had the greatest range of total body mass, which counters my hypothesis as well. I expected neutered cats to be the heaviest, specifically females. Rather the data here shows that neutered lean cats will have a wide variety of body masses. I decided to dig deeper.

To review, again I found that female, neutered obese cats whom underwent energy restriction were *not* heavier than the male cats, both before and after energy restriction. In fact, after undergoing energy restriction, the male cats retain almost the same body mass as their original weights. In contrast, the female cats lost more weight. This was surprising, as I expected the opposite. Following in the same vein, the intact cats are much lighter than the neutered, which makes sense now given the data and does support my hypothesis. I am hesitant to say this is a pattern and that *all* intact cats will be lighter than the neutered, but it certainley is supported by this data as well as other studies. For example, in a study done by Wei et al., they found that when neutered cats have free choice of food access, there is a statistically significant increase of food intake and weight (2014). They also found concentrations of ghrelin increased and adiponectin levels decreased. All of which indicated that there is an inital post-neutering weight gain [@wei2014early]. Thus, while we cannot make hard and fast conclusions that intact cats will always be more lean, there is a lot of data supporting that they will. 

Again, we're finding more obese male cats after neutering than females. But why? I observed Wei's study and looked into ghrelin and adiponectin. Grehlin is a "hunger hormone", which plays a role in "meal initiation" [@klok2007role]. Adiponectin is secreted by the adipose tissue in the endocrine system. Interestingly, it is effected by testosterone easily, thus it is found more in human females than in human males [@robinson2011clinical]. While humans and cats have dissimilar physiologies, cats are still considered a model organism and I hypothesize that there might be connection between the amount of the ghrelin and adiponectin in males than females, causing this different in body mass. However, such an analysis will have to be saved for another time. 

For a moment I would like to go back to my question as to why not only males were more obese, but why the obese neutered males had a harder time losing weight than the females. I'm inclined to think that this speaks to a molecular issue rather than a behavioral one. Male cats might have different lipid retentions than females, or simply their biochemical pathways when it comes to their own metabolisms might be different. While this could be a future direction, I argue that there is much more to the story than simply genetics/heredity/environment at play. Rather, it is feasible that there are many things at play here.  

In sum, we find that male cats are more obese than female cats, even after energy restriction. This is counter to my hypothesis. However, intact cats are still more lean than neutered, which supports my hypothesis that being neutered will cause a cat to become heavier. Next, I look into how the bacteria within the cat's gut plays a role.

**Bacterial phylas Proteobacteria, Tenericutes, and N/A do not show up in intact cats**

I considered the role of the gut microbiota playing a role in obesity because of its role previously studied in dogs. In a study done by Kieler et al., they found that the composition of the gut microbiome in dogs played an effective role in weight loss. Admittedly, dogs and cats differ. However, the process of a dog's gut microbiome playing *a* role has possibility to identify a link for cats as well. To begin, I did not find any bacterial phyla contending only for male ctas or only for female cats. I only found differences in phyla based on cat phenotype and age. Thus, the findings here can neither support or degrade my hypthesis directly, but it can shed light on future directions which we can better understand the underpinings of neutering and feline obesity.

Firmicutes made an appearance across the lower body mass range and upper body mass range, as shown in Figures 4 and 5. It would be unfair if I didn't admit that the mammalian GI tract is highly diverse [@suchodolski2011intestinal]. Looking for a single community of bacteria possibly affiliated with obesity and neutering is similar to finding a needle in a haystack. But, because Firmicutes did appear most frequently within all body mass sizes, I am inclined to say that the bacterial phyla is *not* playing a direct role on weight or is directly influenced by neutering. 

The overabundance of Firmicutes and Bacteriodetes is consistent with previous studies about feline fecal microbial communities [@ritchie2010characterization]. However, what is not consistent with previous studies about Frimicutes and Bacteriodetes is that, typically, lean mammals will have less Firmicutes and more Bacteriodetes. However, in the data, we found more Firmicutes and less Bacteriodetes in lean neutered cats versus obese neutered cats [@fischer2017effects]. Fischer et al. found this as well and suggest that it speaks to a difference on the Class, Family, or Order level rather than broad Phylum level. This suggests a future analysis of the different classes or families of Firmicutes and Bacteriodetes between lean neutered and obese neutered cats. 

What perhaps is more proactive is not looking at the bacteria phyla which appeared in both males, females, and phenotypes, but observing the ones which only occured in a few. We saw that all neutered cats, both lean and obese, had three extra bacterial phylas than the intact cats in Table 3 and Figure 8. The three bacterial phyla were Proteobacteria, Tenericutes, and a mysterious N/A. While I cannot say what the non-applicable is, although I hypothesize that it might be due to poor sequencing or an alternative error, I can explore Proteobacteria and Tenericutes. Proteobacteria, interestingly, is the phyla most commonly found in healthy cats in a study done by Older et al. in 2017 when observing healthy cats to allergic cats. This was confusing, as I now do not understand why a bacterial phyla commonly found in healthy cats would not exist in the intact cats. This is where I circled back to my main points on *why* neutering is so important. Neutering is done not only as population control, but it also prevents disease and keeps felines healthy (unless obesity is involved). Thus, this is the only reason why I imagine Proteobacteria wouldn't show up in the intact cats whereas they normally should. Tenericutes, on the other hand, are often found in the placenta microbiome but have many roles. I did not find anything explicity about Tenericutes, but more research can always be done.

In sum, we find that Firmicutes are probably not playing a role in my hypothesis. However, I learned that the two bacterial phyla (Bacteriodetes and Firmicutes) have contrasting relationships versus the human version, indiciating that more research must be done on smaller levels. The findings are important becuase when it comes to domesticated cats, we end up sharing microbiomes with our furry friends [@song2013cohabiting]. Knowing about their microbiome can tell us a lot about ourselves as well. Lastly, while I didn't find anything explicity connecting the microbiome to my hypothesis that female cats should be larger, I still have age to look at.

**From our data, age does not play an explicit role**

Age has been shown to alter the microbiota community within humans [@mariat2009firmicutes]. So, it prompted me to ask if that the same for felines? If so, would it also be affected by neutering and weight gain/maintenance? If it played a role in my hypothesis, then either getting neutered at a younger or older age will contribute to weight gain and either females or males will be effected the most. As background, typically felines undergo neutering early in life [@deusch2015longitudinal]. In a study done by Deusch et al., they wanted to identify the connection between feline fecal microbiome and age, diet, enviornment, and gender. Stopping sexual development before achieving reproductive age has, hypothetically, some affect. They found that only age of being neutered had a statistical association to alterations in fecal microbiomes and thus on overall feline health.

I found that within the data, neutered and obese females were the oldest (Table 2). This indicates to me that males are neutered at an earlier age, which as mentioned before can alter the production of certain hormones leading to weight gain. Noticeably however, lean intact cats have no differences between the males and females (Figure 7). Why would intact show no difference but neutered would show such a drastic one? This might be sample design. Arguably, if I were to do the experiment, I would attempt to get cats all within a more narrow, similar age range. While it would imply that age plays a more specific role, we could atleast narrow down if age plays a role in weight gain when neutered. 

To go back to Deusch et al.'s findings, I examined age to bacterial phylas. In Table 3 we saw females are still, on average, older. I'm hesitant to use any of my findings in the analysis to make a distinct relationship between age, sex, neutering, and overall body mass. My reason is that due to the sample design, the ages of all cats are too broad of a range: for the lean intact it was 1-10 years, for the lean neutered it was 4-12 years, and for the obese neutered it was a staggering 1-11 years. Any of the cats whom were castrated or spayed had done so within a large window of 1-6 years. The broad ranges of the above all lead me to dismiss the findings of age given the study design. However, for the future I would indicate similar spay and castration times as well as ages to better regulate/understand the effects of limiting certain sex hormones on weight gain. The answer to which might lead to a foundational answer to my original question.

In sum, I dismiss age as a factor only due to the experimental data and design. I contend that, given another design with more narrow ranges, I could learn more about my original hypothesis.

**Alternative Conclusions/Future Directions**

If the weight gain caused by neutering is not attributed by the sex of the cat, as my hypothesis precits and was disproven, what else could? Underlying my hypothesis was the presumption that the male and female hormones were explicit. However, it might be the hormones working in tandem with other pathways. In a study done by Hoenig et al., they found that "changes in NEFA suppression, caloric intake, and leptin concentrations may be indicators of, and possible risk factors for, the development of obesity in cats after neutering" [@hoenig2002effects]. As also indicated previously, ghrelin and adiponectin levels are shown to have some influence on the mass of humans, dogs, and felines. All of which desire further research.

As mentioned above, the abnormal concentrations of Firmicutes and Bacteriodetes compared to what we would normally expect desire further research. I would argue this is less of a priorty than other avenues, as the bacterial phyla show up in many studies of many different types of cats. However, if we looked at the phylas and their classes showing up in cats of specific sex, age, body mass, and time of neutering, we would perhaps get a better idea of how the gut microbiome directly effects or is directly effected by neutering. 

In addition, metabolic rate might differ among the cats, even on the level of sex. Weight loss would become more difficult if the metabolic rate was different among cats of either certain sex or age of neutering [@fettman1997effects].

Of course, there are factors which play a part that cannot be studied on a molecular level. These factors are breed, owner-pet relationship, and the owner's perception of cat's body size. Not many studies have been done on the difficulty to lose weight based on a feline's breed, however that could be a future direction. 

Lastly, my data also has a lot of errors, causing me to worry as to whether it so greatly influenced my results as to give me an improper conclusion. For this reason, I must contend that while the data from how much I could parse out shows a certain answer, the confluence of improperly done sequencing techniques might have inevitably biased my answers. In addition, the sample design didn't include an obese intact phenotype, which could have told a different story. The way they designed it can cause a lot of bias. 

**Final Thoughts**

From my analysis, I conclude that there is a lot of research which still needs to be done on neutering, weight gain, and the underlying reaosns as to *why* neutering has this effect. Females appear to not be larger than males, despite my earlier hypothesis, and I did not find a direct influence from the gut microbiome and age. Thus, if I were to continue, I would pick a more narrow range of study subjects and study the sex hormone levels of the felines right after neutering and to be continually monitored. While we joke about fat cats and their cuteness, the health risks attributed by obesity are severe. And if we are to be true about our love for our furry friends, despite their often times weird way of reciprocating, the least we can do as human scientists is find out how to best keep them alive while maintaining population control. 

# Sources Cited
