---
title: "Final Report: The Effects of Neutering on Female and Male Adult Cats"
author: "Chelsea Noack"
date: "November 29, 2017"
output:
    github_document: default
    word_document: default
bibliography: references.bib
csl: bioinformatics.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
REFERENCE: 2-3 pages will be 1000-1500

The glorification of fat cats, cats which are overweight or obese, is an attraction rather than a concern in our culture. For what we often mistake as chubby cuteness is a danger to our furry friends. Cats whom are obese are at risk for diabetes type 2, lameness, oral cavity disease, urinary tract disease, neoplasia, and more [@takosova2016feline]. In addition, we find that obesity is one of the most predominant nutritional disorders within cats, staggeringly "between 11.5% and 63% [of] cats [being] overweight or obese" [@tarkosova2016feline]. The causes of obesity are, unfortunately, difficult to pinpoint. Diet is a factor, as well as genetics, environment, and overall physiologies. But is neutering a cause? Previous studies indicate that neutering may also play a factor into the weight gain of cats [@kanchuk2003weight] and they found specifically that it was an increased intake of energy leading to weight gain in neutered (male) cats, rather than less energy expenditure.

Neutering itself is proven to be important for cat population control. Feral cats pose many problems for the human population and conservation, as well as their own health. There is an estimated 50-80 million feral cats in the United States and reportedly 80% of kittens are born each year from feral cats [@ireland2016spatial]. They're known to be an invasive species given the amount of wildlife they kill, as they are generalist predators. For example on islands they are reportedly responsible for contributing to 14% of bird, mammal, and reptile extinctions and 8% of critically endangered birds, mammals, and reptiles [@medina2011global]. In addition to conservation issues, feral cats can also infect us with diseases such as rabies and toxoplasmosis [@ireland2016spatial]. Thus the topic becomes complicated. While population control is a serious issue not just for feline health but for conservation and human health, neutering as an effective/popular method may also lead to obesity, which in turn could still lack benefit for the cat. Finding out more about neutering and its effects on the feline can lead to a better understanding of the true advantages to neutering, or inspire alternative methods for population control. 

I use data from the study titled *Effects of obesity, energy restriction and neutering on the faecal microbiota of cats* by Manuela M. Fischer, et al. in order to better understand the implications behind neutering and cat weight. 

Hypothesis: 
1. Being neutered will increase body mass. 
2. Females will be more obese, most significantly after neutering. This is based off of what we know about human females and how their sex hormones, specifically estrogen, can cause weight gain despite lifestyle [@grantham2014estrogen].
3. To further support my analysis, I will also look at the gut microbiome in the cats and age of the cats. I argue that they will not play as significant of a role as the sex of the cat, especially after being neutered, due to sex hormones. 

"Feline obesity is associated with a significant increase in FT4 (free serum throxine) within the normal range" [@ferguson2007obesity]. 

# Methods

## Study design

Fischer et al. used twenty-four cats from the University of California, Davis with a median age of 6.4 years. All cats were pathogen free domestic shorthairs. From the twenty-four cats, eight of the cats were obese with four males and four females. Eight cats were lean and not neutered, again with four males and four females. Eight cats, six males and two females, were lean and neutered [@fischer2017effects]. All twenty-four cats had the same diet before and during the study. Every cat underwent a physical and blood sample before the experiment began.

Cats which were lean and both neutered or not neutered housed together. All had the same diet *but* their intake was not measured. The night before the fecal sampling, the cats slept alone. Fischer et al. performed BCS (body composition scores) for each cat before sampling. 

The obese, neutered cats underwent castration/spaying 1-6 years before the study [@fischer2017effects]. Unlike the lean cats, the obese cats slept alone. The obese neutered ate the normal diet given to all cats for ten days. For six weeks, they ate 60-70% of the prior intake. Again, unlike the lean cats, their intake was now measured. Fischer et al. recorded the cats' weights every week. They also recorded fecal samples before and after the dietary restriction began [@fischer2017effects].

## Sample origin and sequencing

Before sequencing the data, Fischer et al. collected fecal samples from the cat litter box once a day for three days straight. They monitored the cats and they did not collect any samples which were 15 minutes old. To get the bacterial DNA, Fischer et al. used bead-beating methods. Bead-beating is a way to release DNA, RNA, and proteins in the cells of a sample. They used a commercial DNA extraction kit [@fischer2017effects]. To bead-beat, samples go in tubes then on a homogenizer. A homogenizer which is a lab tool used to mix materials using high energy. They mixed "for 60 seconds at a speed of 4 m/s" [@fischer2017effects].

The 16S rRNA amplified using a universal bacterial primer. A universal bacterial primers is important to interpret PCR results. Especially for microbial communities [@mao2012coverage]. PCR (polymerase chain reaction)  is a technique used to make a lot of copies of a segment of DNA. 

After PCR amplification, Fischer et al. added amplicon products of different samples. They purified the samples using AgencourtAmpure beads. The beads were important for high-throughput purification of the PCR amplicons. Then the samples combined in equal concentrations. Fischer et al. sequenced th samples using Roche 454 FLX titanium instruments [@fischer2017effects]. Pyrosequencing using instruments like Roche 454 have big advantages versus traditional Sanger sequencing. Primarily because they can examine many samples from many microbial communities [@tamaki2011analysis]. Once the raw sequences were complete, I downloaded them for my analysis. I downloaded them from NCBI’s Sequence Read Archive (http://www.ncbi.nlm.nih.gov/Traces/sra) under accession no. SRP066010. 

## Computational

First, I downloaded the data from the provided link. This was a two-part process. First, I had to download the SRAtoolkit from NCBI's website. Because the data I used was in .sra format, the toolkit converts it to .fastq. I wanted .fastq so I could perform a quality check later on. After downloading the toolkit, I performed a check to make sure it worked. The steps taken came from the toolkit's website (https://www.ncbi.nlm.nih.gov/books/NBK158900/). Once I assured that it worked, I wrote a Bash script which downloaded the data into my computer. Because the files were fairly small, this did not take much time. The files downloaded into my raw data folder and saved as fastq, thanks to the toolkit. 

With the data downloaded into my computer, I transferred it to RStudio to begin my analysis. In order to do so, I downloaded necessary, but helpful, libraries. The libraries included dplyr, tidyr, knitr, ggplot2, citr, seqinr, mctoolsr, dada2, and phyloseq. Dplyr, tidyr, knitr, and ggplot2 are general use packages to help analyze the data. 

DADA2 and phyloseq were extremely important for this process. DADA2 comes from bioconductor. Bioconductor serves as the primary package to clean up and analyze our data. DADA2 specifically cleans up amplicon errors [@callahan2016]. It can also extract sample names from their original .fastq names. Phyloseq takes on microbiome analysis in R [@mcmurdie2013]. Once all files downloaded and organized, I did a QC (quality check) report on each individual file. 

Initially, I trimmed the files based off of only the QC scores (they dropped around ~400 bp). After later reviewing the amount of errors in the sequences, I decided to extend the trim to 600. To counterbalance this, I made sure the quality cut off was at a smaller number (originally at 5, now at 2). Thus, we would get more data to analyze with adequate quality. After a proper trim and error check, I deleted any duplicate sequences. 454 Roche technology sequenced the data, so I used the suggested parameters based off of (https://benjjneb.github.io/dada2/). This is when DADA2 proved to be vital, as it facilitated the process for my specific amplicon data set. In the final steps, I observed the distribution of trimmed and denoised sequences. While the data was poorly sequenced, for my purposes it would work alright. Then, I removed chimeras. 

Now, the data given is fully processed. Before analysis, I created a metadata table and a melted table to observe during research. 

# Results

I began my analysis using a grid plot in Figure 1 of the overall abundance of the four sample groups between males and females. In the figure, we see that males who are neutered are in the highest abundance, followed by males who are obese undergoing weight loss and females who are obese undergoing weight loss. In the least abundance is males who are intact, obese males, and females who are intact. Given this information, I went forward and created a density plot of the overall abundance of total mass between the four sample groups. While this did not include sex, it indicated that, generally, cats obese undergoing weight loss were the most dense. The second-most dense were the obese cats. The least dense were intact cats whom kept a generally uniform distribution of total mass. That cats who were neutered were neither most or least dense, but showed the greatest range of total body mass (Figure 2).

**The role of body mass**

After analyzing the abundance of different phenotypes by sex and the distribution of body masses, I wanted to look deeper into the body mass question. What were the mean body masses among the four phenotypes for both male and female cats? In Table 1, the phenotype with the highest mean body mass are the obese undergoing weight loss female cats, followed by obese male cats undergoing weight loss. The phenotype with the lowest mean body mass are the intact male cats, followed by intact females. The medians of body mass are obese male cats and neutered female cats. Given the information in Table 1, I wanted to see it visually. Figure 3 depicts Table 1, and in it we see that the difference between females and males in terms of body mass to phenotype is not very significant. The most difference is between intact females and males. The obese cats undergoing weight loss are nearly similar between males and females. We see the general trend of increased body mass and phenotype along both sexes with males being generally larger. 

**The role of microbial communities**

Next, I wanted to see which microbial communities were playing a role, if at all. Thus, I made a density plot indicating the abundance of certain microbial communities in different phenotypes and sexes. Figure 4 indicates the lower range of body masses from the data, while Figure 5 shows the upper range. I divided these graphs for better visualization of the data as there were many data points to take into account. In Figure 4, we see that female cats dominate the lower weight range class among all phenotypes, showing a fairly normal distribution. Firmicutes shows the highest abundance in the lower weight ranges of both males and females. In Figure 5, males dominate the upper body weight range with uneven peaks, while females have a moderate range with only two dominant peaks. Firmicutes, again, have the highest density of both males and females. Visibly, Bacteriodetes also have a high density between the two sexes. Proteobacteria and Fusobacteria have intense fluctuations between sex and body mass. 

The microbial community question continued to linger, so I dug further. In Figure 6, I used a grid plot instead for better visualization of my question. We see a new phylum option as 'not applicable' having the highest abundance of female and male cats. After, Firmicutes, Bacteriodetes, and Actinobacteria. The lowest abundance is Fusobacteria, Proteobacteria, and Tenericutes.

**The role of age**

Lastly, I wanted to see if age played a role between the sexes when it came to remaining obese. In Figure 7, we use the data from Table 2 to see the mean ages across the phenotypes. In Table 2, we see the neutered females, and obese (with or without weight loss) females are the oldest. The neutered males, intact males, and obese males are the youngest. In Figure 7, we see females in all phenotypes have the larger mean age groups while males dominate the younger age groups. The intact cats have similar mean age patterns between males and females. The neutered, obese, and obese undergoing weight loss phenotypes show the strongest variance between the sexes. Next, I wanted to find out the mean sex across the bacteria phylas and cat phenotypes for females and males. This would be a far more hefty task. In Table 3, we see that females are still, on average, older. Intact phenotypes have considerably less bacteria phyla indicated here than the other phenotypes. Females on average are still older than males. I wanted to visualize the data in the best way possible, so a bar graph seemed like an effective solution. In Figure 8, we see the intact phenotype has three fewer bacterial phyla than the other phenotypes. Neutered phenotypes have older ages, on average. Obese and obese undergoing weight loss typically have similar ages as well as bacterial phylas. 

```{r load-libraries, echo=FALSE, message = FALSE}
# Be sure to install these packages before running this script
# They can be installed either with the intall.packages() function
# or with the 'Packages' pane in RStudio

# load general-use packages
library("dplyr")
library("tidyr")
library("knitr")
library("ggplot2")

# this package allows for the easy inclusion of literature citations in our Rmd
# more info here: https://github.com/crsh/citr
# and here:
# http://rmarkdown.rstudio.com/authoring_bibliographies_and_citations.html
library("citr")

# These are the primary packages well use to clean and analyze the data
# this package needs to be installed from bioconductor -- it's not on CRAN
# see info here: https://benjjneb.github.io/dada2/dada-installation.html
library("dada2")

# This to export a fasta of our final denoised sequence variants
library("seqinr")

# To install this you have to install from GitHub
# See more info here: https://github.com/leffj/mctoolsr
# run this -- install.packages("devtools")
# and then this -- devtools::install_github("leffj/mctoolsr")
library("mctoolsr")

# And this to visualize our results
# it also needs to be installed from bioconductor
library("phyloseq")

# Install devtools package
devtools::install_github("ropenscilabs/gramr")
```

```{r}
load("output/phyloseqobject.RData")
```

```{r meltedtable, echo=FALSE}
# Compile into a table
melted_obj <- psmelt(phyloseq_obj)
```

```{r gridplot-sex-abundance-phenotype}
# First figure indicating the general schema of the data
ggplot(data=melted_obj, 
          aes(x=host_sex_s, y=Abundance)) +
  geom_point() +
  facet_grid(. ~ host_phenotype_s) +
  labs(title = "Abundance of male and female cats with different phenotypes",
       x = "Sex of Cat")
```

**Figure 1:** A grid plot of abundance of female and male cats whom are intact, neutered, obese, and who were obese before weight loss. 

```{r density-mass-phenotype}
# Second figure using density to show the range of body mass to phenotype
melted_obj %>%
ggplot(aes(host_tot_mass_s, fill = host_phenotype_s)) +
  geom_density(alpha = 0.1) +
  xlim(3400, 7000) +
  labs(title = "Distribution of body size in different phenotypes",
       x = "Total Mass of Cats", 
       y = "Density", 
       fill = "Cat Phenotype")
```

**Figure 2**: A density plot of host's total mass to its phenotype. 

```{r make-table-1, echo = FALSE}
# Comment here
meanmass_8 <- melted_obj %>%
  group_by(host_phenotype_s, host_sex_s) %>%
  summarize(mean_bodymass = mean(host_tot_mass_s)) %>%
  arrange(desc(host_phenotype_s)) %>%
  head(n = 8)

# then we can use the `kable()` function to make a nicely formatted
# markdown table
meanmass_8 %>%
  kable()
```

**Table 1**: Mean body mass among phenotypes and sexes. 

```{r}
meanmass_8 %>%
  ggplot(aes(x = host_sex_s,
             y = mean_bodymass,
             fill = host_phenotype_s)) +
    geom_col(position = "dodge") +
  ggtitle("Cat sex to mean body mass over four phenotypes") +
  xlab("Cat Sex") + ylab("Mean Body Mass") + labs(fill = "Cat Phenotype")
```

**Figure 3**: Using the table, the mean body mass visualized with a bar graph. 

```{r density-sex-lowermass-phylum}
# Comment here
melted_obj %>%
ggplot(aes(host_tot_mass_s, fill = host_sex_s, colour = Phylum)) +
  geom_density(alpha = 0.1) +
  xlim(2500, 5000) +
  ggtitle("Lower distribution of body mass across bacterial phyla and cat sex") +
  xlab("Cat total mass") + ylab("Density") + labs(fill = "Cat sex", colour = "Bacteria phylum")
```

**Figure 4**: A density plot of the lower range weight class of females and males and the correlating Phylums. 

```{r density-sex-uppermass-pylum}
# Comment here
melted_obj %>%
ggplot(aes(host_tot_mass_s, fill = host_sex_s, colour = Phylum)) +
  geom_density(alpha = 0.1) +
  xlim(5500, 7000) +
    ggtitle("Upper distribution of body mass across bacterial phyla and cat sex") +
  xlab("Cat total mass") + ylab("Density") + labs(fill = "Cat sex", colour = "Bacteria phylum")
```

**Figure 5**: A density plot of the upper range weight class of females and males and the correlating Phylums. 

```{r gridplot-abundance-sex-phylum}
# Comment here
ggplot(data=melted_obj, 
          aes(x=host_sex_s, y=Abundance)) +
  geom_point() +
  facet_grid(. ~ Phylum) +
  labs(title = "Abundance of male and female cats to bacterial phylum",
       x = "Sex of Cat")
```

**Figure 6**: A grid plot of the abundance of bacterial phylums in both males and females. 

```{r make-table-2, echo = FALSE}
# Comment here
meanage_10 <- melted_obj %>%
  group_by(host_phenotype_s, host_sex_s) %>%
  summarize(mean_age = mean(host_age_s)) %>%
  arrange(desc(host_phenotype_s)) %>%
  head(n = 10)

# then we can use the `kable()` function to make a nicely formatted
# markdown table
meanage_10 %>%
  kable()
```

**Table 2**: A table of the mean ages for each phenotype in male and female cats. 

```{r baplot-phenotype-sex-age, echo = FALSE}
# Comment here
meanage_10 %>%
  ggplot(aes(x = host_phenotype_s,
             y = mean_age,
             fill = host_sex_s)) +
    geom_col(position = "dodge") +
  labs(title = "Male and female phenotypes measured to mean age",
       x = "Cat Phenotype",
       y = "Cat Age",
       fill = "Cat Sex")
```

**Figure 7**: A bar plot indicating different mean ages between the categories intact, neutered, obese, and obese after weight loss among females and males. 

```{r make-table-3, echo = FALSE}
# Comment here
meanaps_30 <- melted_obj %>%
  group_by(host_phenotype_s, Phylum, host_sex_s) %>%
  summarize(mean_agephylumsex = mean(host_age_s)) %>%
  arrange(desc(host_phenotype_s)) %>%
  head(n = 50)

# then we can use the `kable()` function to make a nicely formatted
# markdown table
meanaps_30 %>%
  kable()
```

**Table 3**: A table indicating mean ages for the different cat phenotypes, bacteria phylas, and sex. 

```{r barplot-age-phenotype-phylum, echo = FALSE}
# Comment here
meanaps_30 %>%
  ggplot(aes(x = host_phenotype_s,
             y = mean_agephylumsex,
             fill = Phylum)) +
    geom_col(position = "dodge") +
    labs(title = "Phenotypes and bacterial phyla measured to mean age",
       x = "Cat Phenotype",
       y = "Cat Age",
       fill = "Phylum")
```

**Figure 8**: A bar plot showing age against phylum and category of intact, neutered, obese, and obese after weight loss. 

# Discussion

My hypothesis was that female cats are more obese than male cats, especially after being neutered. However, the data indicates opposite results. Male cats are heavier than female cats within all phenotypes; intact, neutered, obese, and obese with energy restriction (Table 1). The findings are supported by supplementary studies indicating that male cats often gain weight after being neutered. Specifically, neutering male cats in the earlier stages (between weeks 3-10) or in the late stages (weeks 34-41) of sexual development can alter feeding behaviors [@allaway2016metabolic].

First, I contend that there is a slight skew in the data when it comes to the abundance of male cats versus female cats, as shown in Figure 1. Arguably, the skew is too slight to make a difference. However, clearly there is some reason as to why male cats are heavier than the female cats. Thus, rather than using the sex to determine abundance, using the density of a particular phenotype to total body mass shows a different story. 

**Male cats are more obese than female cats**

We found that those who were obese undergoing weight loss had the thickest density of sequences in our data. I found this interesting as there were more lean cats, both intact and neutered, than the eight obese cats in total. In fact, the intact lean cats generally had a uniform distribution of total mass. Neutered lean cats had the greatest range of total body mass, which counters my hypothesis as well. I expected neutered cats to be the heaviest, specifically females. Rather the data here shows that neutered lean cats will have a wide variety of body masses. I decided to dig deeper.

To review, again I found that female, neutered obese cats whom underwent energy restriction were *not* heavier than the male cats, both before and after energy restriction. In fact, after undergoing energy restriction, the male cats retain almost the same body mass as their original weights. In contrast, the female cats lost more weight. This was surprising, as I expected the opposite. Following in the same vein, the intact cats are much lighter than the neutered, which makes sense now given the data and does support my hypothesis. I am hesitant to say this is a pattern and that *all* intact cats will be lighter than the neutered, but it certainley is supported by this data as well as other studies. For example, in a study done by Wei et al., they found that when neutered cats have free choice of food access, there is a statistically significant increase of food intake and weight (2014). They also found concentrations of ghrelin increased and adiponectin levels decreased. All of which indicated that there is an inital post-neutering weight gain [@wei2014early]. Thus, while we cannot make hard and fast conclusions that intact cats will always be more lean, there is a lot of data supporting that they will. 

Again, we're finding more obese male cats after neutering than females. But why? I observed Wei's study and looked into ghrelin and adiponectin. Grehlin is a "hunger hormone", which plays a role in "meal initiation" [@klok2007role]. Adiponectin is secreted by the adipose tissue in the endocrine system. Interestingly, it is effected by testosterone easily, thus it is found more in human females than in human males [@robinson2011clinical]. While humans and cats have dissimilar physiologies, cats are still considered a model organism and I hypothesize that there might be connection between the amount of the ghrelin and adiponectin in males than females, causing this different in body mass. However, such an analysis will have to be saved for another time. 

For a moment I would like to go back to my question as to why not only males were more obese, but why the obese neutered males had a harder time losing weight than the females. I'm inclined to think that this speaks to a molecular issue rather than a behavioral one. Male cats might have different lipid retentions than females, or simply their biochemical pathways when it comes to their own metabolisms might be different. While this could be a future direction, I argue that there is much more to the story than simply genetics/heredity/environment at play. Rather, it is feasible that there are many things at play here.  

In sum, we find that male cats are more obese than female cats, even after energy restriction. This is counter to my hypothesis. However, intact cats are still more lean than neutered, which supports my hypothesis that being neutered will cause a cat to become heavier. Next, I look into how the bacteria within the cat's gut plays a role.

**Bacterial phylas Proteobacteria, Tenericutes, and N/A do not show up in intact cats**

I considered the role of the gut microbiota playing a role in obesity because of its role previously studied in dogs. In a study done by Kieler et al., they found that the composition of the gut microbiome in dogs played an effective role in weight loss. Admittedly, dogs and cats differ. However, the process of a dog's gut microbiome playing *a* role has possibility to identify a link for cats as well. To begin, I did not find any bacterial phyla contending only for male ctas or only for female cats. I only found differences in phyla based on cat phenotype and age. Thus, the findings here can neither support or degrade my hypthesis directly, but it can shed light on future directions which we can better understand the underpinings of neutering and feline obesity.

Firmicutes made an appearance across the lower body mass range and upper body mass range, as shown in Figures 4 and 5. It would be unfair if I didn't admit that the mammalian GI tract is highly diverse [@suchodolski2011intestinal]. Looking for a single community of bacteria possibly affiliated with obesity and neutering is similar to finding a needle in a haystack. But, because Firmicutes did appear most frequently within all body mass sizes, I am inclined to say that the bacterial phyla is *not* playing a direct role on weight or is directly influenced by neutering. 

The overabundance of Firmicutes and Bacteriodetes is consistent with previous studies about feline fecal microbial communities [@ritchie2010characterization]. However, what is not consistent with previous studies about Frimicutes and Bacteriodetes is that, typically, lean mammals will have less Firmicutes and more Bacteriodetes. However, in the data, we found more Firmicutes and less Bacteriodetes in lean neutered cats versus obese neutered cats [@fischer2017effects]. Fischer et al. found this as well and suggest that it speaks to a difference on the Class, Family, or Order level rather than broad Phylum level. This suggests a future analysis of the different classes or families of Firmicutes and Bacteriodetes between lean neutered and obese neutered cats. 

What perhaps is more proactive is not looking at the bacteria phyla which appeared in both males, females, and phenotypes, but observing the ones which only occured in a few. We saw that all neutered cats, both lean and obese, had three extra bacterial phylas than the intact cats in Table 3 and Figure 8. The three bacterial phyla were Proteobacteria, Tenericutes, and a mysterious N/A. While I cannot say what the non-applicable is, although I hypothesize that it might be due to poor sequencing or an alternative error, I can explore Proteobacteria and Tenericutes. Proteobacteria, interestingly, is the phyla most commonly found in healthy cats in a study done by Older et al. in 2017 when observing healthy cats to allergic cats. This was confusing, as I now do not understand why a bacterial phyla commonly found in healthy cats would not exist in the intact cats. This is where I circled back to my main points on *why* neutering is so important. Neutering is done not only as population control, but it also prevents disease and keeps felines healthy (unless obesity is involved). Thus, this is the only reason why I imagine Proteobacteria wouldn't show up in the intact cats whereas they normally should. Tenericutes, on the other hand, are often found in the placenta microbiome but have many roles. I did not find anything explicity about Tenericutes, but more research can always be done.

In sum, we find that Firmicutes are probably not playing a role in my hypothesis. However, I learned that the two bacterial phyla (Bacteriodetes and Firmicutes) have contrasting relationships versus the human version, indiciating that more research must be done on smaller levels. The findings are important becuase when it comes to domesticated cats, we end up sharing microbiomes with our furry friends [@song2013cohabiting]. Knowing about their microbiome can tell us a lot about ourselves as well. Lastly, while I didn't find anything explicity connecting the microbiome to my hypothesis that female cats should be larger, I still have age to look at.

**From our data, age does not play an explicit role**

Age has been shown to alter the microbiota community within humans [@mariat2009firmicutes]. So, it prompted me to ask if that the same for felines? If so, would it also be affected by neutering and weight gain/maintenance? If it played a role in my hypothesis, then either getting neutered at a younger or older age will contribute to weight gain and either females or males will be effected the most. As background, typically felines undergo neutering early in life [@deusch2015longitudinal]. In a study done by Deusch et al., they wanted to identify the connection between feline fecal microbiome and age, diet, enviornment, and gender. Stopping sexual development before achieving reproductive age has, hypothetically, some affect. They found that only age of being neutered had a statistical association to alterations in fecal microbiomes and thus on overall feline health.

I found that within the data, neutered and obese females were the oldest (Table 2). This indicates to me that males are neutered at an earlier age, which as mentioned before can alter the production of certain hormones leading to weight gain. Noticeably however, lean intact cats have no differences between the males and females (Figure 7). Why would intact show no difference but neutered would show such a drastic one? This might be sample design. Arguably, if I were to do the experiment, I would attempt to get cats all within a more narrow, similar age range. While it would imply that age plays a more specific role, we could atleast narrow down if age plays a role in weight gain when neutered. 

To go back to Deusch et al.'s findings, I examined age to bacterial phylas. In Table 3 we saw females are still, on average, older. I'm hesitant to use any of my findings in the analysis to make a distinct relationship between age, sex, neutering, and overall body mass. My reason is that due to the sample design, the ages of all cats are too broad of a range: for the lean intact it was 1-10 years, for the lean neutered it was 4-12 years, and for the obese neutered it was a staggering 1-11 years. Any of the cats whom were castrated or spayed had done so within a large window of 1-6 years. The broad ranges of the above all lead me to dismiss the findings of age given the study design. However, for the future I would indicate similar spay and castration times as well as ages to better regulate/understand the effects of limiting certain sex hormones on weight gain. The answer to which might lead to a foundational answer to my original question.

In sum, I dismiss age as a factor only due to the experimental data and design. I contend that, given another design with more narrow ranges, I could learn more about my original hypothesis.

**Alternative Conclusions/Future Directions**

If the weight gain caused by neutering is not attributed by the sex of the cat, as my hypothesis precits and was disproven, what else could? Underlying my hypothesis was the presumption that the male and female hormones were explicit. However, it might be the hormones working in tandem with other pathways. In a study done by Hoenig et al., they found that "changes in NEFA suppression, caloric intake, and leptin concentrations may be indicators of, and possible risk factors for, the development of obesity in cats after neutering" [@hoenig2002effects]. As also indicated previously, ghrelin and adiponectin levels are shown to have some influence on the mass of humans, dogs, and felines. All of which desire further research.

As mentioned above, the abnormal concentrations of Firmicutes and Bacteriodetes compared to what we would normally expect desire further research. I would argue this is less of a priorty than other avenues, as the bacterial phyla show up in many studies of many different types of cats. However, if we looked at the phylas and their classes showing up in cats of specific sex, age, body mass, and time of neutering, we would perhaps get a better idea of how the gut microbiome directly effects or is directly effected by neutering. 

In addition, metabolic rate might differ among the cats, even on the level of sex. Weight loss would become more difficult if the metabolic rate was different among cats of either certain sex or age of neutering [@fettman1997effects].

Of course, there are factors which play a part that cannot be studied on a molecular level. These factors are breed, owner-pet relationship, and the owner's perception of cat's body size. Not many studies have been done on the difficulty to lose weight based on a feline's breed, however that could be a future direction. 

Lastly, my data also has a lot of errors, causing me to worry as to whether it so greatly influenced my results as to give me an improper conclusion. For this reason, I must contend that while the data from how much I could parse out shows a certain answer, the confluence of improperly done sequencing techniques might have inevitably biased my answers. In addition, the sample design didn't include an obese intact phenotype, which could have told a different story. The way they designed it can cause a lot of bias. 

**Final Thoughts**

From my analysis, I conclude that there is a lot of research which still needs to be done on neutering, weight gain, and the underlying reaosns as to *why* neutering has this effect. Females appear to not be larger than males, despite my earlier hypothesis, and I did not find a direct influence from the gut microbiome and age. Thus, if I were to continue, I would pick a more narrow range of study subjects and study the sex hormone levels of the felines right after neutering and to be continually monitored. While we joke about fat cats and their cuteness, the health risks attributed by obesity are severe. And if we are to be true about our love for our furry friends, despite their often times weird way of reciprocating, the least we can do as human scientists is find out how to best keep them alive while maintaining population control. 

# Sources Cited
